Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-05-11 src/tex/ms.tex
update

The IMRPhenomD model is constructed by combining three individually fitted parts
into one coherent waveform model, which consists of the inspiral, intermediate, and merger-ringdown part, 
\kw{Check if the following equation is correct. I think it is off by some smoothing factors.} 
\kl{I don't remember there is any smoothing factors. They are just connected using step functions, then the entire waveform is made to have continuous first derivative by fixing some of the coefficients for the two connections.}
\begin{align}\label{eq:joint_waveform}
	h(f,\theta,\Lambda^i)=h_{\mathrm{ins}}(f,\theta,\Lambda^i) + h_{\mathrm{int}}(f,\theta,\Lambda^i) + h_{\mathrm{rd}}(f,\theta,\Lambda^i).
\end{align}

Instead of fitting the strain, which is a highly oscillatory function that is
difficult to fit, the amplitude and phase are fitted since they are smoother
functions. In each part, the amplitude and phase are made using simple functions
of frequency such as polynomials or lorentzians. Specifically, the merger-ringdown 
amplitude is fitted by a lorentzian and the other parts are fitted using polynomials. 
\begin{equation}\label{eq:amplitude}
\begin{aligned}
	A_0&\equiv\sqrt{\frac{2\eta}{3\pi^{1/3}}}f^{-7/6}\\
	A_{\mathrm{ins}}(f;\theta)&=A_{\mathrm{PN}}(f;\theta)+A_0\sum_{i=1}^3\rho_if^{(6+i)/3} \\
	A_{\mathrm{int}}(f;\theta)&=A_0(\delta_0+\delta_1f+\delta_2f^2+\delta_3f^3+\delta_4f^4) \\
	A_{\mathrm{rd}}(f;\theta)&=A_0\left[\gamma_1\frac{\gamma_3f_{\mathrm{damp}}}{(f-f_{\mathrm{RD}})^2+(\gamma_3f_{\mathrm{damp}})^2}e^{-\frac{\gamma_2(f-f_{\mathrm{RD}})}{\gamma_3f_{\mathrm{damp}}}}\right],  
\end{aligned}
\end{equation}
where $A_{\mathrm{PN}}$ is the post-newtonian expansion of the insprial amplitude up to order $A_0f^2$, $f_{\mathrm{damp}}$ is the damping frequency, and $f_{\mathrm{RD}}$ is the frequency at ringdown. The ansatzes for phase can be found in \citep{khan2016frequency}. These simple analytic functions
consists of parameters $\Lambda^i=\{\rho_i,\delta_i,\gamma_i\}$, which are defined as follows. 
\begin{align} \label{eq:Lambda}
	\Lambda^i&=\lambda_{00}^i+\lambda_{10}^i\eta \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)(\lambda_{01}^i+\lambda_{11}^i\eta+\lambda_{21}^i\eta^2) \nonumber \\ 
	&+(\chi_{\mathrm{PN}}-1)^2(\lambda_{02}^i+\lambda_{12}^i\eta+\lambda_{22}^i\eta^2) \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)^3(\lambda_{03}^i+\lambda_{13}^i\eta+\lambda_{23}^i\eta^2),
\end{align}
where $\lambda$ are fitting coefficients obtained during calibration, $\eta$ is
the symmetric mass ratio, and $\chi_{\mathrm{PN}}$ is the post-Newtonian spin
parameter, which is defined as 
\begin{align}
	\chi_{\mathrm{PN}}=\frac{m_1\chi_1+m_2\chi_2}{m_1+m_2}-\frac{38\eta}{113}(\chi_1+\chi_2).
\end{align}
Here, $m_{1,2}$ and $\chi_{1,2}$ are the primary and secondary mass and spin,
respectively. Finally, the individual segments are first connected directly using 
step functions. Then, by fixing coefficients in the intermediate segment, one can make 
the final waveform is continuous in its first derivative.

@@

In order to recalibrate the model, we need to define an objective function that
quantify the performance of the model. A common choice for such metric is the
\textit{mismatch} function \citep{husa2016frequency}. It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveform we are comparing, and $t_0$ and $\phi_0$
are time shift and phase shift respectively. The $\langle h_1, h_2 \rangle$ is
commonly referred as the inner product, which is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density(PSD) of noise from the instrument, $f_{\mathrm{max}}$ and $f_{\mathrm{min}}$ are
the relevant maximum and minimum frequencies for the integration. The mismatch
is a quantity that is closely related to the mean square error (MSE) between the
two waveforms, as it is often considered a weighted version of the MSE.

\kl{The maximum is calculated by comparing the phase of IMRPhenomD and NR waveforms. We know that they are off by a linear relation of $f$, 
\begin{align*}
	\phi_{NR}-\phi_{IMR}=(2\pi t_0)f+\phi_0.
\end{align*}
Since I have $\phi_{NR}$, $\phi_{IMR}$ and the frequency array, I did a linear regression to get $2\pi t_0$ and $\phi_0$.}

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.~\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table~\ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the bottom-right 
region. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch has additional improvement due to most training waveforms being
equal-spin. In the top-left region, the improvement is significant, except for a
few defects caused by some testing waveforms having $q>4$. This can be seen clearer in 
Fig.~\ref{fig:all_quadrants}, which we see that the optimized histogram shifts downward uniformly. 
The optimized waveforms in the bottom-right region have a higher mismatch than the original ones,
indicating poor fitting. These waveforms correspond to the waveforms in the high mismatch tail 
in Fig.~\ref{fig:q148}, which we can see in Fig.~\ref{fig:all_quadrants}. Thus, we show that 
further optimization in a smaller subspace does not improve the results.

>>>>

The IMRPhenomD model is constructed by combining three individually fitted parts
into one coherent waveform model, which consists of the inspiral, intermediate, and merger-ringdown part, 
\begin{align}\label{eq:joint_waveform}
	h(f,\theta,\Lambda^i)=h_{\mathrm{ins}}(f,\theta,\Lambda^i) + h_{\mathrm{int}}(f,\theta,\Lambda^i) + h_{\mathrm{rd}}(f,\theta,\Lambda^i).
\end{align}
Instead of fitting the strain, which is a highly oscillatory function that is
difficult to fit, the amplitude and phase are fitted since they are smoother
functions. In each part, the amplitude and phase are made using simple functions
of frequency such as polynomials or lorentzians. Specifically, the merger-ringdown 
amplitude is fitted by a lorentzian and the other parts are fitted using polynomials. 
\begin{equation}\label{eq:amplitude}
\begin{aligned}
	A_0&\equiv\sqrt{\frac{2\eta}{3\pi^{1/3}}}f^{-7/6}\\
	A_{\mathrm{ins}}(f;\theta)&=A_{\mathrm{PN}}(f;\theta)+A_0\sum_{i=1}^3\rho_if^{(6+i)/3} \\
	A_{\mathrm{int}}(f;\theta)&=A_0(\delta_0+\delta_1f+\delta_2f^2+\delta_3f^3+\delta_4f^4) \\
	A_{\mathrm{rd}}(f;\theta)&=A_0\left[\gamma_1\frac{\gamma_3f_{\mathrm{damp}}}{(f-f_{\mathrm{RD}})^2+(\gamma_3f_{\mathrm{damp}})^2}e^{-\frac{\gamma_2(f-f_{\mathrm{RD}})}{\gamma_3f_{\mathrm{damp}}}}\right],  
\end{aligned}
\end{equation}
where $A_{\mathrm{PN}}$ is the post-newtonian expansion of the insprial amplitude up to order $A_0f^2$, $f_{\mathrm{damp}}$ is the damping frequency, and $f_{\mathrm{RD}}$ is the frequency at ringdown. The ansatzes for phase can be found in \citep{khan2016frequency}. These simple analytic functions
consists of parameters $\Lambda^i=\{\rho_i,\delta_i,\gamma_i\}$, which are defined as follows. 
\begin{align} \label{eq:Lambda}
	\Lambda^i&=\lambda_{00}^i+\lambda_{10}^i\eta \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)(\lambda_{01}^i+\lambda_{11}^i\eta+\lambda_{21}^i\eta^2) \nonumber \\ 
	&+(\chi_{\mathrm{PN}}-1)^2(\lambda_{02}^i+\lambda_{12}^i\eta+\lambda_{22}^i\eta^2) \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)^3(\lambda_{03}^i+\lambda_{13}^i\eta+\lambda_{23}^i\eta^2),
\end{align}
where $\lambda$ are fitting coefficients obtained during calibration, $\eta$ is
the symmetric mass ratio, and $\chi_{\mathrm{PN}}$ is the post-Newtonian spin
parameter, which is defined as 
\begin{align}
	\chi_{\mathrm{PN}}=\frac{m_1\chi_1+m_2\chi_2}{m_1+m_2}-\frac{38\eta}{113}(\chi_1+\chi_2).
\end{align}
Here, $m_{1,2}$ and $\chi_{1,2}$ are the primary and secondary mass and spin,
respectively. Finally, the individual segments are first connected directly using 
step functions. Then, by fixing coefficients in the intermediate segment, one can make 
the final waveform is continuous in its first derivative.

@@

In order to recalibrate the model, we need to define an objective function that
quantify the performance of the model. A common choice for such metric is the
\textit{mismatch} function \citep{husa2016frequency}. It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveform we are comparing, and $t_0$ and $\phi_0$
are time shift and phase shift respectively. The $\langle h_1, h_2 \rangle$ is
commonly referred as the inner product, which is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density(PSD) of noise from the instrument, $f_{\mathrm{max}}$ and $f_{\mathrm{min}}$ are
the relevant maximum and minimum frequencies for the integration. $t_0$ and $\phi_0$ are maximized by performing regression between $h_1$ and $h_2$. The mismatch
is a quantity that is closely related to the mean square error (MSE) between the
two waveforms, as it is often considered a weighted version of the MSE.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.~\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table~\ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the bottom-right 
region. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch has additional improvement due to most training waveforms being
equal-spin. In the top-left region, the improvement is significant, except for a
few defects caused by some testing waveforms having $q>4$. This can be seen clearer in 
Fig.~\ref{fig:all_quadrants}, which we see that the optimized histogram shifts downward uniformly. 
The optimized waveforms in the bottom-right region have a higher mismatch than the original ones,
indicating poor fitting. These waveforms correspond to the waveforms in the high mismatch tail 
in Fig.~\ref{fig:q148}, which we can see in Fig.~\ref{fig:all_quadrants}. Thus, we show that 
further optimization in a smaller subspace does not improve the results. \kw{I am a bit confused about why you can get worse result.}
